---
title: "Data visualization and exploration - QPCR data"
author: "Kaitlyn"
date: "2019-04-11"
output: md_document
editor_options:
  chunk_output_type: console
---

Set working directory and load data.

```{r setup, message = F, warning = F}
library(here)
library(tidyverse)

# import cleaned data file
genotypes <- read_csv(here::here("data", "cleaned-triplicates-genotypes-qpcr.csv"))
```

## qPCR by storage method
```{r concentration by storage}
ggplot(genotypes, aes(x = stor, y = extracted_conc, fill = stor)) +
  geom_boxplot() +
  ylab("DNA Concentration") +
  xlab("Storage Method") +
  ggtitle("DNA Concentration Across Storage Methods")

ggplot(genotypes, aes(x = stor, y = log(extracted_conc), fill = stor)) +
  geom_boxplot() +
  ylab("DNA Concentration (log)") +
  xlab("Storage Method") +
  ggtitle("DNA Concentration Across Storage Methods")
```

## qPCR vs condition, by storage method
```{r concentration by storage * condition}
ggplot(genotypes, aes(x = cond, y = log(extracted_conc), col = stor)) +
  geom_jitter() +
  facet_wrap(~ stor) +
  ylab("DNA Concentration (log)") +
  xlab("Sample Condition") 
```

Calculate some summary stats
```{r summary stats}
qpcr.summary <- genotypes %>%
  group_by(stor) %>%
  summarize(mean = mean(extracted_conc, na.rm = TRUE),
            median = median(extracted_conc, na.rm = TRUE),
            sd = sd(extracted_conc, na.rm = TRUE),
            min = min(extracted_conc, na.rm = TRUE),
            max = max(extracted_conc, na.rm = TRUE)
            )

qpcr.summary
```

Quick ANOVA to see if there is a difference in DNA concentration across storage methods (there isn't).

```{r anova comparison}
fit <- aov(extracted_conc ~ stor, data = genotypes)
summary(fit)
```


## Relationship between genotyping and qPCR

Here, plotting the number of amplifying genotypes vs. DNA concentration (qPCR). It doesn't look like there is much of a relationship, even though we might expect a positive relationship.
```{r genotyping vs qpcr}
ggplot(genotypes, aes(log(extracted_conc), n, col = stor)) +
  geom_jitter()
```
